~\Large Stickle!

### 2023-05-12 manuscript/paper.Rmd
numbering
Update paper.

# Models  {#sec:models}
## Imputation model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.  Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.  In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.

@@

## OU model 
After imputing sex, we fit an OU model for different phenotypes of interest.  WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.

@@

\begin{align}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right. 
\end{align}

In this setting, for $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as

$$
\begin{aligned}
u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\end{aligned}
$$
$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.  In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.

@@

$$
\begin{aligned}
u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\end{aligned}
$$
This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.

Priors: 
$$
\begin{aligned}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\end{aligned}
$$

>>>>

# Models  {#sec:models}
## Imputation Model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.  Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.  In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.

@@

## Completed Data Model 
For each of the $M = 100$ imputed datasets, we fit an OU model for different phenotypes of interest.  We choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.

@@

\begin{align}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:x}
\end{align}

Look at \ref{eq:x}

For $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as

\begin{align}
u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\label{eq:ar1}
\end{align}

$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.  In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.

@@

\begin{align}
u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\label{eq:u1}
\end{align}

This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.

Priors:

\begin{align}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\label{eq:priors}
\end{align}

### 2023-05-12 manuscript/paper.tex
Update Paper
Update paper.

\title{~\Large Stickle!}
\author{\large Matthew Stuart \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Akhil Ghosh \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Yoel Stuart \vspace{-1.1mm}\\
\normalsize Department of Biology \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Center for Data Science and Consulting; Department of
Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

\[
\begin{aligned}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
\] In this setting, for \(g \in \{f,m\}\), \(\theta_g\), is the defined
as the overall mean of the phenotype of interest for each gender, and
\(u_{gt}\) are the gender specific random effects, measuring the
difference between the gender and time-specific mean and the overall
gender-specific mean, which we fit to an autoregressive process of lag
1, which is the discretization of an OU process. For \(t = 2,\ldots,T\)
and \(g \in \{f,m\}\), the AR1 model for is defined as

\[
\begin{aligned}
u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\end{aligned}
\] \(\kappa_g\) represents the gender-specific correlation between the
previous and current time periods' random effects. In addition,
\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
\(h \in \mathbb{N}\).

@@

\[
\begin{aligned}
u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\end{aligned}
\] This model choice is to make the distribution of the random effects
stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
\(s,t \in \{1,\dots,T\}\).

Priors: \[
\begin{aligned}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\end{aligned}
\]

>>>>

\title{~\Large Stickle!}
\author{\large Matthew Stuart \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:mstuart1@luc.edu}{\texttt{mstuart1@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Akhil Ghosh \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Yoel E. Stuart \vspace{-1.1mm}\\
\normalsize Department of Biology \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Center for Data Science and Consulting; Department of
Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

\begin{align}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\label{eq:x}
\end{align}

Look at \ref{eq:x}

For \(g \in \{f,m\}\), \(\theta_g\), is the defined as the overall mean
of the phenotype of interest for each gender, and \(u_{gt}\) are the
gender specific random effects, measuring the difference between the
gender and time-specific mean and the overall gender-specific mean,
which we fit to an autoregressive process of lag 1, which is the
discretization of an OU process. For \(t = 2,\ldots,T\) and
\(g \in \{f,m\}\), the AR1 model for is defined as

\begin{align}
u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\label{eq:ar1}
\end{align}

\(\kappa_g\) represents the gender-specific correlation between the
previous and current time periods' random effects. In addition,
\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
\(h \in \mathbb{N}\).

@@

\begin{align}
u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\label{eq:u1}
\end{align}

This model choice is to make the distribution of the random effects
stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
\(s,t \in \{1,\dots,T\}\).

Priors:

\begin{align}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \nonumber \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \nonumber \\
\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \nonumber \\
\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\label{eq:priors}
\end{align}
